---
title: |
       | \pkg{RcppCNPy}: 
       | Reading and writing \pkg{NumPy} binary files

author:
  - name: Dirk Eddelbuettel
    affiliation: a
  - name: Wush Wu
    affiliation: b
address:
  - code: a
    address: \url{http://dirk.eddelbuettel.com}
  - code: b
    address: \url{https://github.com/wush978}

lead_author_surname: Eddelbuettel and Wu

doi: "https://cran.r-project.org/package=RcppCNPy"

abstract: |
  This vignette introduces the \pkg{RcppCNPy} package for reading
  and writing files created by or for the \pkg{NumPy} module for
  \proglang{Python}.

acknowledgements: |
  This short paper about the \pkg{RcppCNPy} package can be cited as \citet{JOSS:RcppCNPy}; 
  see the `citation("RcppCNPy")` command in R for details.

footer_contents: "RcppCNPy Vignette"

output: pinp::pinp

header-includes:
  \newcommand{\proglang}[1]{\textsf{#1}}
  \newcommand{\pkg}[1]{\textbf{#1}}
  \newcommand{\R}{\proglang{R}\ } 
  \newcommand{\Rns}{\proglang{R}} 

include-after: |
  \begin{thebibliography}{4}
  \newcommand{\enquote}[1]{``#1''}
  \providecommand{\natexlab}[1]{#1}
  \providecommand{\url}[1]{\texttt{#1}}
  \providecommand{\urlprefix}{URL }
  \expandafter\ifx\csname urlstyle\endcsname\relax
    \providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}\else
    \providecommand{\doi}{doi:\discretionary{}{}{}\begingroup
    \urlstyle{rm}\Url}\fi
  \providecommand{\eprint}[2][]{\url{#2}}
  
  \bibitem[{Eddelbuettel(2013)}]{Eddelbuettel:2013:Rcpp}
  Eddelbuettel D (2013).
  \newblock \emph{Seamless R and C++ Integration with Rcpp}.
  \newblock Use R! Springer, New York.
  \newblock ISBN 978-1-4614-6867-7.
  
  \bibitem[{Eddelbuettel and Fran\c{c}ois(2011)}]{JSS:Rcpp}
  Eddelbuettel D, Fran\c{c}ois R (2011).
  \newblock \enquote{{Rcpp}: Seamless {R} and {C++} Integration.}
  \newblock \emph{Journal of Statistical Software}, \textbf{40}(8), 1--18.
  \newblock \urlprefix\url{http://www.jstatsoft.org/v40/i08/}.
  
  \bibitem[{Eddelbuettel \emph{et~al.}(2017)Eddelbuettel, Fran\c{c}ois, Allaire,
    Ushey, Kou, Russel, Chambers, and Bates}]{CRAN:Rcpp}
  Eddelbuettel D, Fran\c{c}ois R, Allaire J, Ushey K, Kou Q, Russel N, Chambers
    J, Bates D (2017).
  \newblock \emph{{Rcpp}: Seamless {R} and {C++} Integration}.
  \newblock R package version 0.12.12,
    \urlprefix\url{http://CRAN.R-Project.org/package=Rcpp}.
  
  \bibitem[{Eddelbuettel and Wu(2016)}]{JOSS:RcppCNPy}
  Eddelbuettel D, Wu W (2016).
  \newblock \enquote{{RcppCNPy}: Read-Write Support for {NumPy} Files in R.}
  \newblock \emph{The Journal of Open Source Software}, \textbf{1}(5).
  \newblock \doi{10.21105/joss.00055}.
  \newblock \urlprefix\url{https://doi.org/10.21105/joss.00055}.

  \end{thebibliography}


vignette: >
  %\VignetteIndexEntry{RcppCNPy-intro}
  %\VignetteKeywords{Python, NumPy, R, data transfer}
  %\VignettePackage{RcppCNPy}
  %\VignetteEngine{knitr::rmarkdown}
---


```{r echo = FALSE, message = FALSE}
knitr::opts_chunk$set(python.reticulate = FALSE)
if (identical(Sys.info()[['sysname']], "Windows")) {
    knitr::opts_chunk$set(eval = FALSE)                                      
    msg <- paste("Note: Some examples in this vignette require Python", 
                 "but you are running this vignette on Windows where Python",
                 "is much less likely to be present, or even known to be",
                 "missing (i.e. win-builder) so examples will not be evaluated.")
    msg <- paste(strwrap(msg), collapse="\n")
    message(msg) 
}
```

# Motivation

\proglang{Python}\footnote{\url{http://www.python.org}} is a widely-used
and popular programming language. It is deployed in use cases ranging from simple
scripting to larger-scale application development. \proglang{Python} is also popular
for quantitative and scientific application due to the existence of extension
modules such as \pkg{NumPy}\footnote{\url{http://numpy.scipy.org/}} (which is
shorthand for Numeric Python) and many other packages for data analysis.

\pkg{NumPy} is used to efficiently represent $N$-dimensional arrays,
and provides an efficient binary storage model for these files.  In
practice, $N$ is often equal to two, and matrices processed or
generated in \proglang{Python} can be stored in this form. As
\pkg{NumPy} is popular, many project utilize this file format.

\R has no dedicated reading or writing functionality for these type of files.
However, Carl Rogers has provided a small \proglang{Cpp} library called
\pkg{cnpy}\footnote{\url{https://github.com/rogersce/cnpy}} which is released
under the MIT license. Using the `Rcpp modules' feature in
\pkg{Rcpp} \shortcites{CRAN:Rcpp} \citep{JSS:Rcpp,Eddelbuettel:2013:Rcpp,CRAN:Rcpp}, we provide
(some) features of this library to \Rns.

# Examples

## Data creation in \proglang{Python}

The first code example simply creates two files in \proglang{Python}: a
two-dimensional rectangular array as well as a vector.

```{python}
import numpy as np

mat = np.arange(12).reshape(3,4) * 1.1
np.save("fmat.npy", mat)
print mat

vec = np.arange(5) * 1.1
np.save("fvec.npy", vec)
print vec
``` 

As illustrated, \proglang{Python} uses the \proglang{Fortran} convention for storing
matrices and higher-dimensional arrays: a matrix constructed from a single
sequence has its first consecutive elements in its first row---whereas \Rns,
following the \proglang{C} convention, has these first few values in its first
column. This shows that to go back and forth we need to transpose these
matrices (which represented internally as two-dimensional arrays).

## Data reading in \R

We can read the same data in \R using the \code{npyLoad()} function provided
by the \pkg{RcppCNPy} package:

```{r rex1}
library(RcppCNPy)
mat <- npyLoad("fmat.npy")
mat
vec <- npyLoad("fvec.npy")
vec
```

The \proglang{Fortran}-order of the matrix is preserved; we obtain the exact
same data as we stored.

## Reading compressed data in \R

A useful extension to the \pkg{cnpy} library is the support of
\pkg{gzip}-compressed data.

```{r rex2, eval=FALSE}
mat2 <- npyLoad("fmat.npy.gz")
``` 

Support for writing compressed files has been added in version 0.2.0.

## Data writing in \R

Matrices and vectors can be written to files using the \code{npySave()}
function.

```{r rex3}
set.seed(42)
m <- matrix(sort(rnorm(6)), 3, 2)
m
npySave("randmat.npy", m)
v <- seq(10, 12)
v
npySave("simplevec.npy", v)
```

## Data reading in \proglang{Python}

Reading the data back in \proglang{Python} is also straightforward as shown in
the following example:

```{python pyex2}
import numpy as np  
m = np.load("randmat.npy")
print m
v = np.load("simplevec.npy")
print v
```

## Integer support

Support for integer data types has been conditional on use of either
the \code{-std=c++0x} or the \code{-std=c++11} compiler extensions.
Only these standards support the \code{long long int} type needed to
represent \code{int64} data on a 32-bit OS.  Following the release of
\R 3.1.0, it has been enabled by default in \pkg{RcppCNPy} (whereas it
previously required a manual rebuild), and following the release of R
3.3.0 with its updated Windows toolchain, C++11 is now available on
all common R platforms. Consequently, support for large integers in
\pkg{RcppCNPy} is no longer just a compile-time option for some
platforms, but generally available on all (current) R installations.


## Performance

The \R script \code{timing} in the \code{demo/} directory of the package
\pkg{RcppCNPy} provides a simple benchmark.  Given two values $n$ and $k$, a
matrix of size $n \times k$ is created with $n$ rows and $k$ columns. It is
written to temporary files in
i) ascii format using \code{write.table()};
ii) \code{NumPy} format using \code{npySave()}; and
iii) \code{NumPy} format using \code{npySave()} with compression via
the \code{zlib} library (used also by \code{gzip}).

Table~\ref{tab:benchmark} shows some timing comparisons for a matrix with
five million elements.  Reading the \code{npy} data is clearly fastest as it
required only parsing of the header, followed by a single large binary read
(and the transpose required to translate the representation used by \Rns). The
compressed file requires only one-fourth of the disk space, but takes
approximately 2.5 times as long to read as the binary stream has be
transformed.  Lastly, the default ascii reading mode is clearly by far the
slowest.

\begin{table}[bt]
  \begin{center}
    \begin{small}
      \begin{tabular}{rrr}
        \toprule
        {\bf Access method \phantom{X}} & {\bf Time in sec.} & {\bf Relative to best} \\
        \cmidrule(r){1-3}
     \code{npyLoad(pyfile)}   &    0.074 &  1.000  \\
   \code{npyLoad(pygzfile)}   &    0.190 &  2.568 \\
 \code{read.table(}txtfile)   &    4.189 & 56.608 \\
        \bottomrule
      \end{tabular}
    \end{small}
    \caption{Performance comparison of data reads using a matrix of size
      $10^5 \times 50$. File size are 39.7mb for ascii, 40.0mb for npy and
      10.8mb for npy.gz. Ten replications were performed, and total times
      are shown. R 3.3.1 was used on a laptop with an SSD disk. }
    \label{tab:benchmark}
  \end{center}
\end{table}


# Limitations

## Higher-dimensional arrays

\pkg{Rcpp} supports three-dimensional arrays, this could be support in
\pkg{RcppCNPy} as well.

## \code{npz} files

The \pkg{cnpy} library supports reading and writing of sets of arrays; this
feature could also be exported.

# Summary

The \pkg{RcppCNPy} package provides simple reading and writing of
\pkg{NumPy} files, using the \pkg{cnpy} library. Reading of compressed
files is also supported as an extension, offering more compact storage
at the cost of slightly longer read times.

```{r rex4, echo=FALSE}
unlink("fmat.npy")
unlink("fvec.npy")
unlink("randmat.npy")
unlink("simplevec.npy")
```

# Summary

The \pkg{RcppCNPy} package provides simple reading and writing of
\pkg{NumPy} files, using the \pkg{cnpy} library. Reading of compressed
files is also supported as an extension, offering more compact storage
at the cost of slightly longer read times.

<!-- local tweak to the pnasbreak command -->
\renewcommand{\pnasbreak}{\begin{strip}\vskip0pt\end{strip}}
