---
title: Using \pkg{reticulate} to read and write \pkg{NumPy} files

author:
  - name: Dirk Eddelbuettel
    affiliation: a
address:
  - code: a
    address: \url{http://dirk.eddelbuettel.com}

lead_author_surname: Eddelbuettel

doi: "https://cran.r-project.org/package=RcppCNPy"

abstract: |
  This vignette shows how to use the \pkg{reticulate} package for to directly
  access the \pkg{NumPy} module for \proglang{Python}.

footer_contents: "RcppCNPy Vignette"

output: pinp::pinp

header-includes:
  \newcommand{\proglang}[1]{\textsf{#1}}
  \newcommand{\pkg}[1]{\textbf{#1}}
  \newcommand{\R}{\proglang{R}\ } 
  \newcommand{\Rns}{\proglang{R}} 

include-after: |
  \begin{thebibliography}{4}
  \newcommand{\enquote}[1]{``#1''}
  \providecommand{\natexlab}[1]{#1}
  \providecommand{\url}[1]{\texttt{#1}}
  \providecommand{\urlprefix}{URL }
  \expandafter\ifx\csname urlstyle\endcsname\relax
    \providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}\else
    \providecommand{\doi}{doi:\discretionary{}{}{}\begingroup
    \urlstyle{rm}\Url}\fi
  \providecommand{\eprint}[2][]{\url{#2}}

  \bibitem[{Allaire \emph{et~al.}(2018)Allaire, Ushey, RStudio, Tang, Eddelbuettel, 
    Lewis, and Geelnard}]{CRAN:reticulate}
  Allaire JJ, Ushey K, RStudio, Tang Y, Eddelbuettel D, Lewis B, Geelnard M (2018).
  \newblock \emph{{reticulate}: Interface to {Python}}.
  \newblock R package version 1.5,
    \urlprefix\url{http://CRAN.R-Project.org/package=reticulate}.
 
  \bibitem[{Eddelbuettel and Wu(2016)}]{JOSS:RcppCNPy}
  Eddelbuettel D, Wu W (2016).
  \newblock \enquote{{RcppCNPy}: Read-Write Support for {NumPy} Files in R.}
  \newblock \emph{The Journal of Open Source Software}, \textbf{1}(5).
  \newblock \doi{10.21105/joss.00055}.
  \newblock \urlprefix\url{https://doi.org/10.21105/joss.00055}.

  \end{thebibliography}


vignette: >
  %\VignetteIndexEntry{RcppCNPy-reticulate}
  %\VignetteKeywords{Python, NumPy, R, data transfer}
  %\VignettePackage{RcppCNPy}
  %\VignetteEngine{knitr::rmarkdown}
---


```{r echo = FALSE, message = FALSE}
knitr::opts_chunk$set(python.reticulate = FALSE)
knitr::opts_chunk$set(eval = FALSE)  ## fake to not have the packag rely on reticulate
```

# Motivation

The RcppCNPy package by \citet{JOSS:RcppCNPy} provides a simple and
reliable access to NumPy files.  It does not require Python as it relies on
the CNPy library which is connected to R with the help of Rcpp. Now, thanks
to the reticulate package by \citet{CRAN:reticulate}, we can also consider an
alternative which does not require CNPy--but which requires Python.  Thanks
to reticulate, we can (on a correctly set up machine, how to do that is
beyond the scope of this note but described in the reticulate documentation)
use Python to read NumPy data.  And reticulate will faithfully transfer the
data for us.

```{r prep, include=FALSE}
file.copy("~/git/rcppcnpy/tests/fmat.npy", ".")  
file.copy("~/git/rcppcnpy/tests/fmat.npy.gz", ".")  
file.copy("~/git/rcppcnpy/tests/fvec.npy", ".")  
file.copy("~/git/rcppcnpy/tests/imat.npy", ".")  
file.copy("~/git/rcppcnpy/tests/ivec.npy", ".")  
```


# Simple Examples

```{r ex1}
### load reticulate and use it to load numpy
library(reticulate)
np <- import("numpy")

## data reading
(mat <- np$load("fmat.npy"))
```

\begin{ShadedResult}
\begin{verbatim}
#       [,1] [,2] [,3] [,4]
#  [1,]  0.0  1.1  2.2  3.3
#  [2,]  4.4  5.5  6.6  7.7
#  [3,]  8.8  9.9 11.0 12.1
\end{verbatim}
\end{ShadedResult}

```{r ex1a
(vec <- np$load("fvec.npy"))
```

\begin{ShadedResult}
\begin{verbatim}
#  [1] 0.0 1.1 2.2 3.3 4.4
\end{verbatim}
\end{ShadedResult}

Integer data can be read the same way:

```{r ex2}
(imat <- np$load("imat.npy"))
```

\begin{ShadedResult}
\begin{verbatim}
#       [,1] [,2] [,3] [,4]
#  [1,]    0    1    2    3
#  [2,]    4    5    6    7
#  [3,]    8    9   10   11
\end{verbatim}
\end{ShadedResult}

# Compressed Files

The gzip Python module allows us to access compressed files.

```{r ex3}
## compressed data: import gzip
gz <- import("gzip")

## use it to create handle to uncompressed file 
(mat2 <- np$load(gz$GzipFile("fmat.npy.gz","r")))
```

\begin{ShadedResult}
\begin{verbatim}
#       [,1] [,2] [,3] [,4]
#  [1,]  0.0  1.1  2.2  3.3
#  [2,]  4.4  5.5  6.6  7.7
#  [3,]  8.8  9.9 11.0 12.1
\end{verbatim}
\end{ShadedResult}

# Saving Files

Similarly, files can be saved via reticulate access to NumPy.

```{r ex4}
tfile <- tempfile(fileext=".npy")

set.seed(42)
(m <- matrix(sort(rnorm(6)), 3, 2))
```

\begin{ShadedResult}
\begin{verbatim}
#            [,1]     [,2]
#  [1,] -0.564698 0.404268
#  [2,] -0.106125 0.632863
#  [3,]  0.363128 1.370958
\end{verbatim}
\end{ShadedResult}

```{r ex4a}
np$save(tfile, m)

(m2 <- np$load(tfile))
```

\begin{ShadedResult}
\begin{verbatim}
#            [,1]     [,2]
#  [1,] -0.564698 0.404268
#  [2,] -0.106125 0.632863
#  [3,]  0.363128 1.370958
\end{verbatim}
\end{ShadedResult}

```{r ex4b}
all.equal(m, m2)
```

\begin{ShadedResult}
\begin{verbatim}
#  [1] TRUE
\end{verbatim}
\end{ShadedResult}

```{r cleanup, include=FALSE}
unlink("fmat.npy")  
unlink("fmat.npy.gz")  
unlink("fvec.npy")  
unlink("imat.npy")  
unlink("ivec.npy")  
```
